Decoupled Fallback: A Portable Single-Pass GPU Scan

### 2025-06-21 paper.tex
Analysis fixes

As shown in Table~\ref{tab:results}, \emph{DF} achieves at least a $\sim$40\% speedup over \emph{RTS} across all tested devices, closely approaching the theoretically expected $O(3n)/O(2n)$ 50\% speedup, except on the 7900XT and Mali-G78. This result initially appears confusing, given that \emph{DF} outperforms \emph{Memcpy} by 6\% and 13\%, respectively. However, we hypothesized and confirmed a secondary effect of \emph{Chained Scans}: they enforce a queue-like behavior among workgroups. In-flight workgroups with a lower relative rank traverse fewer tiles during their lookbacks, enabling them to exit earlier than higher-ranked workgroups. Because this ordering emerges naturally from differing amounts of computation rather than explicit synchronization, the exit sequence is \emph{k-sorted} or \emph{nearly ordered}, rather than strictly ordered. Scheduling fairness and message-passing visibility can further influence this sequence---for example, an unfair scheduler might allocate more resources to a higher-ranked workgroup, causing it to exit before a lower-ranked counterpart, or a higher-ranked workgroup might observe an updated flag sooner, enabling earlier completion. We confirmed this behavior experimentally by using atomics to record and print the order in which workgroups completed their lookback phase. Unlike a simple \emph{Memcpy} kernel, where all in-flight workgroups begin writing immediately after completing reads, this “ordered exit” pattern facilitates more efficient global memory accesses during the write phase of the scan, as adjacent workgroups are more likely to share memory pages, reducing premature page evictions. We hypothesize that the 7900XT and Mali-G78 have insufficient memory systems to efficiently accommodate all in-flight workgroups, making them particularly sensitive to this effect. To test this hypothesis, we augmented our \emph{Memcpy} kernel with a simplified lookback routine---excluding the scan itself---and observed a performance improvement of approximately 10\% on the 7900XT\@. Thus, the unexpectedly slow baseline \emph{Memcpy} performance on these devices explains why \emph{DF}, despite its strong overall performance, does not fully achieve the theoretical 50\% speedup over \emph{RTS}.

@@

To better understand how devices interact with \emph{DF}, we collect the following statistics during its execution:

\begin{itemize}
  \item \textbf{Spins per Workgroup:} How many times does a workgroup spin while waiting? With perfectly fair scheduling, no waiting is necessary. Since no real-world scheduler is perfectly fair, we want minimal waiting. Despite varying occupancy rates, the average spin rate across tested devices remains low (typically one or fewer), indicating that most workgroups experience minimal delays.

@@

This data reveals two notable behaviors: (1) As demonstrated by the Apple M3, a scheduler can be fair yet still lack FPG, demonstrating that fairness and FPG are not necessarily correlated. (2) The ARM Mali-G78 MP20 exhibits the most unfair scheduling among all tested devices, with a fallback rate of 57.8\% relative to workgroups launched. However, the average number of spins per workgroup is only 2.52---below our fallback trigger threshold of 4 spins. This suggests a bimodal distribution of waiting times, where workgroups either proceed immediately with minimal delay or experience prolonged stalls that trigger fallbacks.

@@

The slight downward trend in lookback lengths observed in Figure~\ref{fig:lookback_length} reflects the shift in dependency resolution strategy as workgroups become increasingly bogged down by fallbacks. Rather than resolving dependencies through lookback traversals, workgroups revert to tile-to-adjacent-tile propagation of readiness information. This emergent behavior, caused by blocking, can be viewed as a computationally enforced variant of \emph{StreamScan}. We connect this behavior to \emph{StreamScan} because we posit that its strictly enforced adjacent-tile dependency propagation yields perfectly ordered workgroup exits, which could further enhance the global write efficiency behavior observed in Section~\ref{sec:SameVsBase}, possibly explaining the slight uptick in performance seen in Figure~\ref{fig:execution_time} on the Mali-G78.

>>>>

As shown in Table~\ref{tab:results}, \emph{DF} achieves at least a $\sim$30\% speedup over \emph{RTS} across all tested devices, closely approaching the theoretically expected $O(3n)/O(2n)$ 50\% speedup on most devices. Most strikingly, \emph{DF} slightly exceeds the performance of \emph{Memcpy} on the 7900XT and Mali-G78, despite performing additional computation. While this result appears counterintuitive—since we expect \emph{Memcpy} to be the upper bound for memory-bound throughput—we attribute this behavior to a secondary effect of \emph{Chained Scans}: they impose a queue-like structure on workgroup progression. Despite entering the chained propagation phase at approximately the same time—subject to scheduler fairness—in-flight workgroups with lower relative rank traverse fewer tiles and therefore exit the phase earlier than higher-ranked workgroups. Because this ordering is computationally rather than logically enforced, the resulting sequence of completions is \emph{k-sorted} or \emph{nearly ordered}, rather than \emph{perfectly ordered}. This emergent ordering is further shaped by scheduler fairness and message-passing latency: for instance, an unfair scheduler might accelerate a higher-ranked workgroup’s completion, or message passing timing might allow a higher-ranked workgroup to observe readiness earlier than its lower-ranked peers.

We confirmed this queue-like progression experimentally by using atomics to record and print the order in which workgroups exited their lookback phase. This 'ordered exit' behavior is notably different from that of a \emph{Memcpy} kernel, where all in-flight workgroups typically begin writing immediately after completing reads, often leading to high-concurrency, uncoordinated global writes. In contrast, \emph{DF}'s staggered exit pattern implicitly throttles the number of concurrent writers, reducing memory pressure and improving locality. Workgroups that exit in near-order are more likely to access adjacent memory regions, reducing page thrashing and avoiding premature eviction of active pages. On devices whose memory systems may be insufficient to support highly concurrent writes---such as the 7900XT and Mali-G78---these effects plausibly contribute to scan throughput that exceeds \emph{Memcpy}, despite the added arithmetic. To test this hypothesis, we augmented our \emph{Memcpy} kernel with a simplified lookback routine---excluding the scan itself---and observed a $\sim$10\% throughput increase on the 7900XT. These findings suggest that \emph{Chained Scan}'s computational structure not only enables progress without explicit synchronization, but also improves global memory efficiency by shaping the temporal pattern of writes, effectively translating computational work into emergent memory coordination.

@@

To better understand how devices interact with \emph{DF}, we collect the following statistics during its execution:
\begin{itemize}
  \item \textbf{Spins per Workgroup:} How many times does a workgroup spin while waiting? With perfectly fair scheduling, no waiting is necessary. Since no real-world scheduler is perfectly fair, we want minimal waiting. Despite varying occupancy rates, the average spin rate across tested devices remains low (typically one or fewer), indicating that most workgroups experience minimal delays.

@@

This data reveals three notable behaviors: (1) As demonstrated by the Apple M3, a scheduler can be fair yet still lack FPG, demonstrating that fairness and FPG are not necessarily correlated. (2) The ARM Mali-G78 MP20 exhibits the most unfair scheduling among all tested devices, with a fallback rate of 57.8\% relative to workgroups launched. However, the average number of spins per workgroup is only 2.52---below our fallback trigger threshold of 4 spins. This suggests a skewed distribution of waiting times, where workgroups either proceed with minimal delay or experience prolonged stalls that trigger fallbacks. (3) Despite this organically unfair scheduling, \emph{Decoupled Fallback} still achieves a 34.6\% speedup over \emph{RTS}, confirming its robustness to scheduler behavior in real-world conditions.

@@

The slight downward trend in lookback lengths observed in Figure~\ref{fig:lookback_length} reflects the shift in dependency resolution strategy as workgroups become increasingly bogged down by fallbacks. Rather than resolving dependencies through lookback traversals, workgroups revert to tile-to-adjacent-tile propagation of readiness information. This emergent behavior, caused by blocking, can be viewed as a computationally enforced variant of \emph{StreamScan}. We connect this behavior to \emph{StreamScan} because we posit that its strictly enforced adjacent-tile dependency propagation yields \emph{perfectly ordered} workgroup exits, which could further enhance the global write efficiency behavior observed in Section~\ref{sec:SameVsBase}, possibly explaining the slight uptick in performance seen in Figure~\ref{fig:execution_time} on the Mali-G78.
